Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer

### 2020-03-24 manuscript/manuscript.tex
accepted some comments, and improved label visibility in figure_4

\documentclass[twocolumn, 11pt]{article}
\usepackage{amsmath,amssymb}
\usepackage[format=hang,labelfont=bf,textfont=small,singlelinecheck=false,justification=justified,margin={20pt,20pt},figurename=Fig.]{caption}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\usepackage{threeparttable}
\usepackage[numbers]{natbib}
\newcommand{\mean}[1]{$\overline{\mbox{#1}}$}
\newcommand{\median}[1]{$\hat{\mbox{#1}}$}

\usepackage{lmodern} % Chloe: better font rendering

@@

\begin{abstract}
Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. \caz{It does so by jointly considering}{Indeed, it allows to jointly consider} the statistical association between gene variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study the utility of six \caz{network methods to discover new biomarkers for breast cancer susceptibility by searching subnetworks with higher association scores to this phenotype}{methods that identify subnetworks with higher association scores to a phenotype to discover new biomarkers for breast cancer susceptibility}. We interrogate a familial breast cancer genome-wide association study (GWAS) focused on \emph{BRCA1/2} negative French women. We perform an in-depth benchmarking of the methods with regards to size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. By trading statistical \caz{astringency}{stringency} for biological meaningfulness, most network methods \caz{get}{give} more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. Interestingly, a combination of solution subnetworks provided a concise subnetwork of 93 genes, enriched in known breast cancer susceptibility genes (\emph{BABAM1}, \emph{BLM}, \emph{CASP8}, \emph{FGFR2}, and \emph{TOX3}, Fisher's exact test P-value~=~$7.8 \times 10^{-5}$) and occupying more central positions in the network than average. Additionally, it shows a general alteration of the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer. Finally, we find a significantly large overlap between the genes in the consensus network and the genes significantly associated in the largest GWAS on susceptibility to breast cancer. In conclusion, this article shows the pertinence of network-based analyses to tackle known issues with GWAS, namely lack of statistical power and of interpretable solutions.
\cazcom{}{Add a sentence to the effect that the computational tools are available on github for ease of application to other data sets?}
\end{abstract}

@@

GWAS have successfully identified thousands of variants underlying many common diseases \cite{buniello_nhgri-ebi_2019}. However, this experimental setting also presents intrinsic challenges. Some of them stem from the high dimensionality of the problem, as every GWAS to date studies more variants than samples are genotyped. This limits the statistical power of the experiment, as only variants with larger effects can be detected \cite{visscher_10_2017}. \caz{And it}{This} is particularly problematic since the prevailing view is that most genetic architectures involve many variants with small effects \cite{visscher_10_2017}. Additionally, to avoid false positives, a conservative multiple test correction is applied, typically the \caz{aforementioned}{previously mentioned} Bonferroni\caz{}{correction}. However, Bonferroni \caz{}{correction} is overly conservative when the statistical tests are correlated, as is the case in GWAS \cite{wang_statistical_2018}. Another open issue is the interpretation of the results, as the functional consequences of most common variants are not well understood. On top of that, recent large-sampled studies suggest that most of the genome contributes to a degree to any complex trait, in accordance with the infinitesimal model \cite{barton_infinitesimal_2017}. The recently proposed omnigenic model \cite{boyle_expanded_2017} offers an explanation: genes are very functionally inter-related and influence each other's behavior, which allows alterations in most genes to impact the subset of ``core'' genes directly involved in the disease's mechanism. Hence, a comprehensive statistical framework which includes the structure of biological data might address the aforementioned issues.

\caz{In this regard}{For this reason,} many authors turn to network biology to handle the complex interplay of biomolecules that lead to disease \cite{furlong_human_2013}. As its name suggests, network biology models biology as a network, where the biomolecules under study, often genes, are nodes, and selected functional relationships are edges that link them. These relationships come from evidence that the genes jointly contribute to a biological function; for instance, their expression is correlated, or their products establish a protein-protein interaction. Under this view, complex diseases are not the consequence of a single altered gene, but of the interaction of multiple interdependent molecules \cite{barabasi_network_2011}. In fact, an examination of biological networks shows that disease genes have differential properties \cite{barabasi_network_2011,pinero_uncovering_2016}. This is particularly true for cancer driver genes, which tend to be key players in connecting different, densely-connected communities of genes. Additionally, as genes that contribute to a disease tend to participate in similar biological functions, studying the neighborhood of known disease genes is effective at identifying new ones \cite{huang_systematic_2018}.

@@

\paragraph{Gene network}
Out of the six methods tested, \cazcom{five use a gene-gene interaction network}{SConES GI does too...} (Section~\ref{methods:methods}). Although their respective statistical frameworks are compatible with any type of network (protein interactions, gene coexpression, regulatory, etc.), we used protein-protein interaction networks (PPIN), as they are interpretable, well characterized, and most of the methods were designed to scale appropriately to them. We built our PPIN from both binary and co-complex interactions stored in the HINT database (release April 2019) \cite{das_hint:_2012}. Unless \caz{specified otherwise}{otherwise specified}, we used only interactions coming from high-throughput experiments, leaving out targeted studies that might bias the topology of the network. Out of the 146\,722 interactions from high-throughput experiments that HINT stores, we were able to map 142\,541 to a pair of gene symbols. The scoring function for the nodes changed from method to method (Section~\ref{methods:methods}).

@@

Genes that contribute to the same function are nearby in the PPIN, might be topologically related to each other in diverse ways (densely interconnected modules, nodes around a hub, a path, etc.). But \caz{accounting for that is not the only choice a developer of a network method needs to make}{this is not the only aspect to model when developing a network method}: how to score the nodes, whether the affected mechanisms form a single connected component or several, how to frame the problem in a computationally efficient fashion, which network to use, etc. Unsurprisingly, multiple solutions have been proposed. We examined six of them: five that explore the PPIN, and one which explores SNP networks. We selected methods that were open source, had an implementation available, and an accessible documentation. Their main differences are summarized in Table~\ref{tab:method_comparison}.

@@

where \emph{k} is the number of genes in $S_i$; z\textsubscript{j} is the Z score of gene $j$, computed as \(\phi\)\textsuperscript{-1}(1 - P-value\textsubscript{j}); and \(\phi\)\textsuperscript{-1} is the inverse normal distribution function.
\item Identify neighboring nodes of \caz{$S$}{$S_i$},  i.e. nodes at distance \(\le\) d. \caz{We set d~=~2.}{}
\item Add the neighboring nodes whose inclusion increases the Z\textsubscript{m+1} more than a threshold Z\textsubscript{m}~\texttimes{}~(1 + r). \caz{In our experiments, we set r~=~0.1.}{}
\item Repeat 2-4 until no further enlargement is possible.
\caz{}{\item Add $S_i$ to the list of subnetworks to return.}
\end{enumerate}
\cazcom{}{Missing here: how to select seed nodes. Is every node a seed node? Every node that has not yet been included in a subgraph?}

Lastly, \caz{the module's}{a subnetwork's} Z-score is normalized as

@@

We used the implementation of dmGWAS in the dmGWAS 3.0 R package \cite{dmgwas}. \caz{}{We used the suggested hyperparameters $d = 2$ and $r = 0.1$} We used the function \emph{simpleChoose} to select the solution subnetwork, which aggregates the top 1\% modules into the solution subnetwork.
\end{description}

\begin{description}
\item[{heinz}] The goal of heinz is to identify the highest-scored connected subgraph on the network \cite{dittrich_identifying_2008}. The authors propose a transformation of the genes' P-value into a score that is negative under no association with the phenotype, and positive when there is. This transformation is achieved by modelling the distribution of P-values by a beta-uniform model (BUM) parameterized by the desired \caz{FDR}{false discovery rate (FDR)}. Thus formulated, the problem is NP-complete, and hence solving it would require a \caz{prohibitely}{prohibitively} long computational time. To solve it efficiently it is re-cast\caz{ed}{} as the Prize-Collecting Steiner Tree Problem (PCST), which seeks to select the connected subnetwork S that maximizes the \emph{profit} p(S), defined as:

@@

were $p(v)~=~w(v) - w'$ is the \emph{profit} of adding a node, \caz{$c(e)~=~w'$}{$c(e)=-w'$} is the \emph{cost} of adding an edge, and $w' = min_{v \in V_{G}} w(v)$ \caz{}{is the smallest node weight of $G$}. All three are positive quantities. Heinz implements the algorithm from \citet{ljubic_algorithmic_2006} which, \caz{although} in practice is often fast and optimal, \caz{}{although} neither is guaranteed. We used BioNet's implementation of heinz \cite{beisser_bionet:_2010,heinz}.

\item[{HotNet2}] HotNet2 was developed to find connected subgraphs of genes frequently mutated in cancer \cite{leiserson_pan-cancer_2015}. To that end, it considers both the local topology of the network and the scores of the nodes. The former is captured by an insulated heat diffusion process: at \caz{the beginning}{initialization}, the score of the node determines its initial heat; iteratively each node yields heat to its ``colder'' neighbors, and receives heat from its ``hotter'' neighbors, while retaining part of its own (hence, \emph{insulated}). This process continues until equilibrium is reached, and results in a \caz{similarity}{diffusion} matrix $F$. $F$ is used to compute the similarity matrix $E$ that \caz{accounts also for similarities in node scores as}{models exchanged heat as}
\begin{equation*} 
E = F \operatorname{diag}(w(V)), 
\end{equation*}
where $\operatorname{diag}(w(V))$ is a diagonal matrix with the node scores in its diagonal. \caz{}{For any two nodes/genes $i$ and $j$, $E_{ij}$ models the amount of heat that diffuses from node $j$ to node $i$, which can be interpreted as a (non-symmetric) similarity between those two nodes.}  To obtain densely connected subnetworks, HotNet2 prunes $E$, only preserving edges such that $w(E) > \delta$. Lastly, HotNet2 evaluates the statistical significance of the subnetworks by comparing their size to the size of networks obtained by permuting the node scores. We \caz{scored the nodes}{assigned initial node scores} as in \citet{nakka_gene_2016}, assigning a score of 0 for the genes with low probability of being associated to the disease, and -log\textsubscript{10}(P-value) to those likely to be. In this dataset, the threshold separating both was a P-value of 0.125, which was obtained using a local FDR approach \cite{scheid_twilight;_2005}. HotNet2 has two parameters: the restart probability \(\beta\), and the threshold heat \(\delta\). Both parameters are set automatically by the algorithm, \caz{and are robust}{which is robust to their values} \cite{leiserson_pan-cancer_2015}. HotNet2 is implemented in Python \cite{hotnet2}.

@@

where \(\lambda\) and \(\eta\) are hyperparameters that control the sparsity and the connectivity of the model. \caz{}{The connectivity term penalizes disconnected solutions, with many edges between a node that is selected and a node that isn't.} Given a $\lambda$ and an $\eta$, the aforementioned problem has a unique solution, that SConES finds using a graph min-cut procedure. As in \citet{azencott_efficient_2013}, we selected \(\lambda\) and \(\eta\) by cross-validation, choosing the values that produce the most stable solution across folds. In this case, the selected hyperparameters were \(\eta\)~=~3.51, \(\lambda\)~=~210.29 for SConES GS; \(\eta\)~=~3.51, \(\lambda\)~=~97.61 for SConES GM; and \(\eta\)~=~3.51, \(\lambda\)~=~45.31 for SConES GI. We used the version on SConES implemented in the R package martini \cite{martini}.
\end{description}

\begin{description}
\item[{SigMod}] SigMod aims at identifying the highest-scoring, most densely connected gene subnetwork \cite{liu_sigmod:_2017}. It addresses an optimization problem similar to that of SConES (Equation \ref{eq:scones}), but \caz{using the adjacency matrix rather than the Laplacian matrix (Section~\ref{methods:notation}), to quantify solutions containing many edges}{the connectivity term encourages connected solutions by favoring solutions where many edges connect two selected nodes, rather than penalizing disconnected ones}.

@@

SigMod presents three important differences with SConES. First it is designed for gene-gene networks. Second, \caz{by replacing the adjacency by the Laplacian matrix,}{} it favors subnetworks containing many edges. SConES, instead, penalizes connections between the selected and unselected nodes. Third, it returns a single connected subnetwork, which it achieves by exploring a grid of hyperparameters and processing their respective solutions. Specifically, for the range of \(\lambda\)~=~\(\lambda\)\textsubscript{min}, \dots{}, \(\lambda\)\textsubscript{max} for the same \(\eta\), it prioritizes the solution with the largest change in size from \(\lambda\)\textsubscript{n} to \(\lambda\)\textsubscript{n+1}. Such a large change implies that the network is densely interconnected. This results in one candidate solution for each \(\eta\), which are processed by removing any node not connected to any other. A score is assigned to each candidate solution by summing their node scores and normalizing by size. The candidate solution with the highest standardized score is the chosen solution. SigMod is implemented in an R package \cite{sigmod}.
\end{description}

@@

We searched for pathways enriched in the gene subnetworks produced by the \caz{aforementioned}{above} methods. We conducted an hypergeometric test on pathways from Reactome \cite{Jassal2019} using the R package ReactomePA \cite{Yu2016}. The universe of genes included any gene that we could map to a SNP in the iCOGS array (Section \ref{methods:node_score}). Pathways with an adjusted P-value \textless~0.05 were deemed significant.

@@

We evaluated multiple properties of the different method through a 5-fold subsampling setting. We applied each method to 5 random subsets of the original dataset containing 80\% of the samples (\emph{train set}). When pertinent, we evaluated the solution on the remaining 20\% (\emph{test set}). We used the 5 repetitions to estimate the average and the \caz{}{standard} deviation of the different measures.

@@

We compared the runtime, the number of selected genes/SNPs, and the stability (sensitivity of the result to small changes in the input\caz{}{, here, using different subsets of the data}) of the different network methods. The stability was measured using \caz{the}{} Pearson correlation between different runs as suggested by \citet{nogueira_measuring_2016}.

@@

Additionally, we used the summary statistics from the Breast Cancer Association Consortium (BCAC), a meta-analysis of case-control studies conducted in multiple countries which included 13\,250\,642 SNPs (after imputation) genotyped on 228\,951 women of European ancestry from the general population \cite{Michailidou2017}. Hence, a high proportion of breast cancer cases investigated in BCAC are sporadic, while GENESIS is a homogeneous dataset focusing on the French high-risk population attending the family cancer clinics. Despite these differences, \caz{there should be}{we expect a} shared genetic architecture, especially at the gene level. For that purpose, we searched for associated genes as in Section~\ref{methods:node_score}. We used all the SNPs to measure gene association, but only those available after quality control on GENESIS for the SNP-level comparisons.

@@

We developed computational pipelines for several steps of GWAS analyses, such as physically mapping SNPs to genes, computing gene scores, and performing six different network analyses. For each of those processes, we created a pipeline with a clear interface that should work on any GWAS dataset. They are compiled in \url{https://github.com/hclimente/gwas-tools}. Although the GENESIS data is not public, the code \caz{that applies these pipelines to it}{to apply the pipelines to this data}, as well as the code that reproduces all the analyses in this article are available at \url{https://github.com/hclimente/genewa}. We deposited all the produced gene subnetworks on NDEx (\url{http://www.ndexbio.org}), under the UUID e9b0e22a-e9b0-11e9-bb65-0ac135e8bacf.

@@

Closer examination reveals two regions (3p24 and 8q24) having low, albeit not genome-wide significant, P-values. Both of them have been associated to breast cancer susceptibility in the past \cite{brisbin_meta-analysis_2011,search_newly_2009}. We applied an L1-penalized logistic regression on the whole dataset (Section~\ref{methods:classifier}), \cazcom{a machine learning algorithm that searches a small subset of SNPs which provide good classification accuracy}{describe it in the methods rather than here? How was the regularization parameter chosen?}. It selected 100 SNPs, both from all aforementioned regions and from other regions (Supplementary Figure~\ref{sfig:snp_gene_manhattan}C). Yet, it is unclear why those SNPs were selected, as emphasized by the high P-value of some of them, which further complicates the biological interpretation. Moreover, and in opposition to what the omnigenic model predicts, the genes to which this SNPs map to (Section~\ref{methods:node_score}) are not interconnected in the PPIN (Section~\ref{methods:networks}). \caz{}{In addition, the classification performance of the method is very low, and L1-penalized methods select only one of several correlated variables and are prone to instability, which further complicates interpretation.} This motivates exploring network methods, which trade statistical significance for biological relevance to find susceptibility subnetworks. In fact, such methods provided comparable classification \caz{accuracies}{performance} to L1-penalized logistic regression \cazcom{(Figure~\ref{fig:benchmark}D)}{Is it B? I don't seel L1-penalized logistic regression on this figure; also with these types of performance I'm not sure it's worth making that point. Maybe frame it as ``do no worse'' than ``have comparable performance''...}, while providing more interpretable solutions.

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[height=.75\textheight]{./figures/figure_1.png}
  \caption{\label{fig:solution_overview} Overview of the solutions produced by the different network methods (Section~\ref{methods:methods}) on the GENESIS dataset. As LEAN did not produce any significant gene (adjusted P-value < 0.05), it was excluded. Unless indicated otherwise, results refer to genes, except for SConES GI which are at the SNP-level. \textbf{(A)} Manhattan plots of SNPs/genes; in black, the method's solution. The Bonferroni threshold is indicated by a red line (2.54 \texttimes{} 10\textsuperscript{-7} for SNPs, 1.53 \texttimes{} 10\textsuperscript{-6} for genes). \textbf{(B)} \cazcom{\caz{Pearson correlation}{Overlap} between the genes selected by each of the methods\caz{}{, measured by Pearson correlation between indicator vectors}.}{You only explain how to interpret Pearson correlation in Figure 3, although it should maybe come in the Methods when you introduce stability} \textbf{(C)} Precision and recall of the evaluated methods with respect to Bonferroni-significant SNPs/genes in GENESIS (left) and BCAC (middle), and breast cancer susceptibility genes (right, only at the gene-level). For reference, we added a gray line with a slope of 1. \textbf{(D)} Solution networks produced by the different methods.}
\end{figure}

>>>>

\documentclass[twocolumn, 11pt]{article}
\usepackage{amsmath,amssymb}
\usepackage[format=hang,labelfont=bf,textfont=small,singlelinecheck=false,justification=justified,margin={20pt,20pt},figurename=Fig.]{caption}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\usepackage{threeparttable}
\usepackage[numbers]{natbib}
\usepackage{lmodern} % Chloe: better font rendering 
\newcommand{\mean}[1]{$\overline{\mbox{#1}}$}
\newcommand{\median}[1]{$\hat{\mbox{#1}}$}

@@

\begin{abstract}
Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. Indeed, it allows to jointly consider the statistical association between gene variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study the utility of six methods that identify subnetworks with higher association scores to a phenotype to discover new biomarkers for breast cancer susceptibility. We interrogate a familial breast cancer genome-wide association study (GWAS) focused on \emph{BRCA1/2} negative French women. We perform an in-depth benchmarking of the methods with regards to size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. By trading statistical stringency for biological meaningfulness, most network methods give more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. Interestingly, a combination of solution subnetworks provided a concise subnetwork of 93 genes, enriched in known breast cancer susceptibility genes (\emph{BABAM1}, \emph{BLM}, \emph{CASP8}, \emph{FGFR2}, and \emph{TOX3}, Fisher's exact test P-value~=~$7.8 \times 10^{-5}$) and occupying more central positions in the network than average. Additionally, it shows a general alteration of the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer. Finally, we find a significantly large overlap between the genes in the consensus network and the genes significantly associated in the largest GWAS on susceptibility to breast cancer. In conclusion, this article shows the pertinence of network-based analyses to tackle known issues with GWAS, namely lack of statistical power and of interpretable solutions.
\cazcom{}{Add a sentence to the effect that the computational tools are available on github for ease of application to other data sets?}
\end{abstract}

@@

GWAS have successfully identified thousands of variants underlying many common diseases \cite{buniello_nhgri-ebi_2019}. However, this experimental setting also presents intrinsic challenges. Some of them stem from the high dimensionality of the problem, as every GWAS to date studies more variants than samples are genotyped. This limits the statistical power of the experiment, as only variants with larger effects can be detected \cite{visscher_10_2017}. This is particularly problematic since the prevailing view is that most genetic architectures involve many variants with small effects \cite{visscher_10_2017}. Additionally, to avoid false positives, a conservative multiple test correction is applied, typically the previously mentioned Bonferroni correction. However, Bonferroni correction is overly conservative when the statistical tests are correlated, as is the case in GWAS \cite{wang_statistical_2018}. Another open issue is the interpretation of the results, as the functional consequences of most common variants are not well understood. On top of that, recent large-sampled studies suggest that most of the genome contributes to a degree to any complex trait, in accordance with the infinitesimal model \cite{barton_infinitesimal_2017}. The recently proposed omnigenic model \cite{boyle_expanded_2017} offers an explanation: genes are very functionally inter-related and influence each other's behavior, which allows alterations in most genes to impact the subset of ``core'' genes directly involved in the disease's mechanism. Hence, a comprehensive statistical framework which includes the structure of biological data might address the aforementioned issues.

For this reason, many authors turn to network biology to handle the complex interplay of biomolecules that lead to disease \cite{furlong_human_2013}. As its name suggests, network biology models biology as a network, where the biomolecules under study, often genes, are nodes, and selected functional relationships are edges that link them. These relationships come from evidence that the genes jointly contribute to a biological function; for instance, their expression is correlated, or their products establish a protein-protein interaction. Under this view, complex diseases are not the consequence of a single altered gene, but of the interaction of multiple interdependent molecules \cite{barabasi_network_2011}. In fact, an examination of biological networks shows that disease genes have differential properties \cite{barabasi_network_2011,pinero_uncovering_2016}. This is particularly true for cancer driver genes, which tend to be key players in connecting different, densely-connected communities of genes. Additionally, as genes that contribute to a disease tend to participate in similar biological functions, studying the neighborhood of known disease genes is effective at identifying new ones \cite{huang_systematic_2018}.

@@

\paragraph{Gene network}
Out of the six methods tested, \cazcom{five use a gene-gene interaction network}{SConES GI does too...} (Section~\ref{methods:methods}). Although their respective statistical frameworks are compatible with any type of network (protein interactions, gene coexpression, regulatory, etc.), we used protein-protein interaction networks (PPIN), as they are interpretable, well characterized, and most of the methods were designed to scale appropriately to them. We built our PPIN from both binary and co-complex interactions stored in the HINT database (release April 2019) \cite{das_hint:_2012}. Unless otherwise specified, we used only interactions coming from high-throughput experiments, leaving out targeted studies that might bias the topology of the network. Out of the 146\,722 interactions from high-throughput experiments that HINT stores, we were able to map 142\,541 to a pair of gene symbols. The scoring function for the nodes changed from method to method (Section~\ref{methods:methods}).

@@

Genes that contribute to the same function are nearby in the PPIN, might be topologically related to each other in diverse ways (densely interconnected modules, nodes around a hub, a path, etc.). But this is not the only aspect to model when developing a network method: how to score the nodes, whether the affected mechanisms form a single connected component or several, how to frame the problem in a computationally efficient fashion, which network to use, etc. Unsurprisingly, multiple solutions have been proposed. We examined six of them: five that explore the PPIN, and one which explores SNP networks. We selected methods that were open source, had an implementation available, and an accessible documentation. Their main differences are summarized in Table~\ref{tab:method_comparison}.

@@

where \emph{k} is the number of genes in $S_i$; z\textsubscript{j} is the Z score of gene $j$, computed as \(\phi\)\textsuperscript{-1}(1 - P-value\textsubscript{j}); and \(\phi\)\textsuperscript{-1} is the inverse normal distribution function.
\item Identify neighboring nodes of \caz{$S$}{$S_i$},  i.e. nodes at distance \(\le\) d.
\item Add the neighboring nodes whose inclusion increases the Z\textsubscript{m+1} more than a threshold Z\textsubscript{m}~\texttimes{}~(1 + r).
\item Repeat 2-4 until no further enlargement is possible.
\caz{}{\item Add $S_i$ to the list of subnetworks to return.}
\end{enumerate}
\cazcom{}{Missing here: how to select seed nodes. Is every node a seed node? Every node that has not yet been included in a subgraph?}

Lastly, a subnetwork's Z-score is normalized as

@@

We used the implementation of dmGWAS in the dmGWAS 3.0 R package \cite{dmgwas}. We used the suggested hyperparameters $d = 2$ and $r = 0.1$. We used the function \emph{simpleChoose} to select the solution subnetwork, which aggregates the top 1\% modules into the solution subnetwork.
\end{description}

\begin{description}
\item[{heinz}] The goal of heinz is to identify the highest-scored connected subgraph on the network \cite{dittrich_identifying_2008}. The authors propose a transformation of the genes' P-value into a score that is negative under no association with the phenotype, and positive when there is. This transformation is achieved by modelling the distribution of P-values by a beta-uniform model (BUM) parameterized by the desired false discovery rate (FDR). Thus formulated, the problem is NP-complete, and hence solving it would require a prohibitively long computational time. To solve it efficiently it is re-cast as the Prize-Collecting Steiner Tree Problem (PCST), which seeks to select the connected subnetwork S that maximizes the \emph{profit} p(S), defined as:

@@

were $p(v)~=~w(v) - w'$ is the \emph{profit} of adding a node, \caz{$c(e)~=~w'$}{$c(e)=-w'$} is the \emph{cost} of adding an edge, and $w' = min_{v \in V_{G}} w(v)$ is the smallest node weight of $G$. All three are positive quantities. Heinz implements the algorithm from \citet{ljubic_algorithmic_2006} which, in practice is often fast and optimal, although neither is guaranteed. We used BioNet's implementation of heinz \cite{beisser_bionet:_2010,heinz}.

\item[{HotNet2}] HotNet2 was developed to find connected subgraphs of genes frequently mutated in cancer \cite{leiserson_pan-cancer_2015}. To that end, it considers both the local topology of the network and the scores of the nodes. The former is captured by an insulated heat diffusion process: at initialization, the score of the node determines its initial heat; iteratively each node yields heat to its ``colder'' neighbors, and receives heat from its ``hotter'' neighbors, while retaining part of its own (hence, \emph{insulated}). This process continues until equilibrium is reached, and results in a diffusion matrix $F$. $F$ is used to compute the similarity matrix $E$ that models exchanged heat as
\begin{equation*} 
E = F \operatorname{diag}(w(V)), 
\end{equation*}
where $\operatorname{diag}(w(V))$ is a diagonal matrix with the node scores in its diagonal. For any two nodes/genes $i$ and $j$, $E_{ij}$ models the amount of heat that diffuses from node $j$ to node $i$, which can be interpreted as a (non-symmetric) similarity between those two nodes. To obtain densely connected subnetworks, HotNet2 prunes $E$, only preserving edges such that $w(E) > \delta$. Lastly, HotNet2 evaluates the statistical significance of the subnetworks by comparing their size to the size of networks obtained by permuting the node scores. We assigned initial node scores as in \citet{nakka_gene_2016}, assigning a score of 0 for the genes with low probability of being associated to the disease, and -log\textsubscript{10}(P-value) to those likely to be. In this dataset, the threshold separating both was a P-value of 0.125, which was obtained using a local FDR approach \cite{scheid_twilight;_2005}. HotNet2 has two parameters: the restart probability \(\beta\), and the threshold heat \(\delta\). Both parameters are set automatically by the algorithm, which is robust to their values \cite{leiserson_pan-cancer_2015}. HotNet2 is implemented in Python \cite{hotnet2}.

@@

where \(\lambda\) and \(\eta\) are hyperparameters that control the sparsity and the connectivity of the model. The connectivity term penalizes disconnected solutions, with many edges between a node that is selected and a node that is not. Given a $\lambda$ and an $\eta$, the aforementioned problem has a unique solution, that SConES finds using a graph min-cut procedure. As in \citet{azencott_efficient_2013}, we selected \(\lambda\) and \(\eta\) by cross-validation, choosing the values that produce the most stable solution across folds. In this case, the selected hyperparameters were \(\eta\)~=~3.51, \(\lambda\)~=~210.29 for SConES GS; \(\eta\)~=~3.51, \(\lambda\)~=~97.61 for SConES GM; and \(\eta\)~=~3.51, \(\lambda\)~=~45.31 for SConES GI. We used the version on SConES implemented in the R package martini \cite{martini}.
\end{description}

\begin{description}
\item[{SigMod}] SigMod aims at identifying the highest-scoring, most densely connected gene subnetwork \cite{liu_sigmod:_2017}. It addresses an optimization problem similar to that of SConES (Equation \ref{eq:scones}), but the connectivity term encourages connected solutions by favoring solutions where many edges connect two selected nodes, rather than penalizing disconnected ones.

@@

SigMod presents three important differences with SConES. First it is designed for gene-gene networks. Second, it favors subnetworks containing many edges. SConES, instead, penalizes connections between the selected and unselected nodes. Third, it returns a single connected subnetwork, which it achieves by exploring a grid of hyperparameters and processing their respective solutions. Specifically, for the range of \(\lambda\)~=~\(\lambda\)\textsubscript{min}, \dots{}, \(\lambda\)\textsubscript{max} for the same \(\eta\), it prioritizes the solution with the largest change in size from \(\lambda\)\textsubscript{n} to \(\lambda\)\textsubscript{n+1}. Such a large change implies that the network is densely interconnected. This results in one candidate solution for each \(\eta\), which are processed by removing any node not connected to any other. A score is assigned to each candidate solution by summing their node scores and normalizing by size. The candidate solution with the highest standardized score is the chosen solution. SigMod is implemented in an R package \cite{sigmod}.
\end{description}

@@

We searched for pathways enriched in the gene subnetworks produced by the above methods. We conducted an hypergeometric test on pathways from Reactome \cite{Jassal2019} using the R package ReactomePA \cite{Yu2016}. The universe of genes included any gene that we could map to a SNP in the iCOGS array (Section \ref{methods:node_score}). Pathways with an adjusted P-value \textless~0.05 were deemed significant.

@@

We evaluated multiple properties of the different method through a 5-fold subsampling setting. We applied each method to 5 random subsets of the original dataset containing 80\% of the samples (\emph{train set}). When pertinent, we evaluated the solution on the remaining 20\% (\emph{test set}). We used the 5 repetitions to estimate the average and the standard deviation of the different measures.

@@

We compared the runtime, the number of selected genes/SNPs, and the stability (sensitivity of the result to small changes in the input, here, using different subsets of the data) of the different network methods. The stability was measured using Pearson correlation between different runs as suggested by \citet{nogueira_measuring_2016}.

@@

Additionally, we used the summary statistics from the Breast Cancer Association Consortium (BCAC), a meta-analysis of case-control studies conducted in multiple countries which included 13\,250\,642 SNPs (after imputation) genotyped on 228\,951 women of European ancestry from the general population \cite{Michailidou2017}. Hence, a high proportion of breast cancer cases investigated in BCAC are sporadic, while GENESIS is a homogeneous dataset focusing on the French high-risk population attending the family cancer clinics. Despite these differences, we expect a shared genetic architecture, especially at the gene level. For that purpose, we searched for associated genes as in Section~\ref{methods:node_score}. We used all the SNPs to measure gene association, but only those available after quality control on GENESIS for the SNP-level comparisons.

@@

We developed computational pipelines for several steps of GWAS analyses, such as physically mapping SNPs to genes, computing gene scores, and performing six different network analyses. For each of those processes, we created a pipeline with a clear interface that should work on any GWAS dataset. They are compiled in \url{https://github.com/hclimente/gwas-tools}. Although the GENESIS data is not public, the code to apply the pipelines to this data, as well as the code that reproduces all the analyses in this article are available at \url{https://github.com/hclimente/genewa}. We deposited all the produced gene subnetworks on NDEx (\url{http://www.ndexbio.org}), under the UUID e9b0e22a-e9b0-11e9-bb65-0ac135e8bacf.

@@

Closer examination reveals two regions (3p24 and 8q24) having low, albeit not genome-wide significant, P-values. Both of them have been associated to breast cancer susceptibility in the past \cite{brisbin_meta-analysis_2011,search_newly_2009}. We applied an L1-penalized logistic regression on the whole dataset (Section~\ref{methods:classifier}), \cazcom{a machine learning algorithm that searches a small subset of SNPs which provide good classification accuracy}{describe it in the methods rather than here? How was the regularization parameter chosen?}. It selected 100 SNPs, both from all aforementioned regions and from other regions (Supplementary Figure~\ref{sfig:snp_gene_manhattan}C). Yet, it is unclear why those SNPs were selected, as emphasized by the high P-value of some of them, which further complicates the biological interpretation. Moreover, and in opposition to what the omnigenic model predicts, the genes to which this SNPs map to (Section~\ref{methods:node_score}) are not interconnected in the PPIN (Section~\ref{methods:networks}). \caz{}{In addition, the classification performance of the method is very low, and L1-penalized methods select only one of several correlated variables and are prone to instability, which further complicates interpretation.} This motivates exploring network methods, which trade statistical significance for biological relevance to find susceptibility subnetworks. In fact, such methods provided comparable classification performance to L1-penalized logistic regression \cazcom{(Figure~\ref{fig:benchmark}B)}{Is it B? I don't seel L1-penalized logistic regression on this figure; also with these types of performance I'm not sure it's worth making that point. Maybe frame it as ``do no worse'' than ``have comparable performance''...}, while providing more interpretable solutions.

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[height=.75\textheight]{./figures/figure_1.png}
  \caption{\label{fig:solution_overview} Overview of the solutions produced by the different network methods (Section~\ref{methods:methods}) on the GENESIS dataset. As LEAN did not produce any significant gene (adjusted P-value < 0.05), it was excluded. Unless indicated otherwise, results refer to genes, except for SConES GI which are at the SNP-level. \textbf{(A)} Manhattan plots of SNPs/genes; in black, the method's solution. The Bonferroni threshold is indicated by a red line (2.54 \texttimes{} 10\textsuperscript{-7} for SNPs, 1.53 \texttimes{} 10\textsuperscript{-6} for genes). \textbf{(B)} \cazcom{Overlap between the genes selected by each of the methods, measured by Pearson correlation between indicator vectors.}{You only explain how to interpret Pearson correlation in Figure 3, although it should maybe come in the Methods when you introduce stability} \textbf{(C)} Precision and recall of the evaluated methods with respect to Bonferroni-significant SNPs/genes in GENESIS (left) and BCAC (middle), and breast cancer susceptibility genes (right, only at the gene-level). For reference, we added a gray line with a slope of 1. \textbf{(D)} Solution networks produced by the different methods.}
\end{figure}
